Co-methylation review

### 2015-05-02 Rmarkdown/appendix.Rmd
Pre-lunch commit. Begin re-write of methsim chapter.
Goknur's suggested changes.

In particular, in the case $n = 2$ we can compute the probability that two
dependent but identically distributed Bernoulli random variables are equal.

@@

>>>>

In particular, in the case $n = 2$ we can compute the probability that two
dependent and identically distributed Bernoulli random variables are equal.

@@

__TODO: Add citations of all packages used in thesis__

### 2015-05-02 Rmarkdown/comethylation_review.Rmd
Goknur's suggested changes.

\begin{chapabstract}
Co-methylation is the dependence structure of DNA methylation data. This dependence structure can be viewed from different perspectives and on different scales, which has led to several disparate definitions of co-methylation. In this chapter I critically review these definitions and put them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}. I highlight various shortcomings of proposed methods, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}. I also review how methods to detect differential methylation have sought to account for and exploit co-methylation.
\end{chapabstract}

@@

Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $Z_{i, h}$ is sampled, that is, $Pr(Z_{h, i} = 1) = p_{h}$. For two distinct CpGs sequenced on two (potentially) distinct amplicons, $Z_{h, i}$ and  $Z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(Z_{h, i} = Z_{h', i'})$. Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from distinct samples with different genome-wide average methylation levels. This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h', i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$. In this setting we can compute the theoretical $PIM^{\ast}$ as follows:

\begin{align*}
PIM^{\ast} &= Pr(Z_{h, i} = Z_{h', i'}) \\
        &= Pr(Z_{h, i} = 0, Z_{h', i'} = 0) + Pr(Z_{h, i} = 1, Z_{h', i'} = 1) \\
        &= Pr(Z_{h', i'} = 0 | Z_{h, i} = 0) Pr(Z_{h, i} = 0) +
           Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(Z_{h, i} = 1) \\
        &= \big [ (1 - p_{h'}) + c (1 - p_{h}) \big ] (1 - p_{h}) +
           \big [ (p_{h'} + c (1 - p_{h})) \big ] p_{h} \\
        &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
\end{align*}

@@

1. $PIM^{\ast}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'}$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
2. $PIM^{\ast}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
3. $PIM^{\ast}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, \allowbreak Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.

@@

The $167$ bp period is "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al. hypothesise that this periodicity is due to histones dictating access to DNA by the DNA methyltransferases.

@@

\citet{Ball:2009kh} use bisulfite padlock probes with $36$ bp single-end reads to profile CpG methylation at approximately $7,000$ CpGs in an Epstein-Barr virus transformed B-lymphocyte cell line. As part of their study, \citeauthor{Ball:2009kh} sought to investigate whether the co-methylation reported by \citet{Eckhardt:2006gh} (Section \label{sec:eckhardt}) occurs at the single-molecule level.

@@

\citet{Landan:2012kp} is notable for its use of the within-read information available from bisulfite-sequencing data. The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs from ultra-deep bisulfite-sequencing (median coverage of each CpG in the target $> 10,000 \times$) of 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time. \citeauthor{Landan:2012kp} also use these data to estimate co-methylation.

As part of their analysis, citeauthor{Landan:2012kp} construct "methylation linkage diagrams" based on a Pearson correlation estimate of within-fragment co-methylation. For each CpG 2-tuple in their targeted regions, they create a $2 \times 2$ contingency table. The general form of the table is shown below in Table \ref{tab:landan}. They then compute the Pearson correlation of each table[^cor_equivalence], $r = \frac{mm_{i, i'} \times uu_{i, i'} - mu_{i, i'} \times um_{i, i'}}{sqrt{u\cdot_{i, i'} \times m\cdot_{i, i'} \times \cdot u_{i, i'} \times m_{i, i'}}}$. Importantly, $r$ is based on within-read measurements from a single sample and so these $r$ are estimates of within-read, within-sample co-methylation.

@@

\begin{table}[h]
\centering
\caption[Counts of reads at a CpG 2-tuple]{Summary of read counts at a CpG 2-tuple, $(i, i'), i \neq i'$. A dot ($\cdot$) indicates that this index is summed over, e.g., $u\cdot_{i, i + 1} = uu_{i, i + 1} + um_{i, i + 1}$.}
\label{tab:landan}
\begin{tabular}{@{}cccc@{}}
\cmidrule(l){2-4}
 & \multicolumn{3}{c}{Second CpG} \\
 \cmidrule(l){2-4}
\multirow{3}{*}{First CpG} & & & \\
 & $uu_{i, i'}$       & $um_{i,i'}$       & $u\cdot_{i, i'}$ \\
 & $mu_{i, i'}$       & $mm_{i, i'}$       & $m\cdot_{i, i'}$ \\
 & $\cdot u_{i, i'}$  & $\cdot m_{i, i'}$  & $d_{i, i'}$      \\ \cmidrule(l){2-4}
\end{tabular}
\end{table}

@@

\begin{table}[h]
\centering
\caption[Notions of 'subject' and 'population' in methylation linkage]{Notions of 'subject' and 'population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}. NB 'person' could be any organism, e.g, 'mouse', 'dog', etc.}
\label{tab:linkage}
\begin{tabular}{@{}lll@{}}
\toprule
                    & Subject               & Population \\ \midrule
Genetic linkage     & Person                   & Population \\
Methylation linkage & (Haploid) genome of cell & Person     \\ \bottomrule
\end{tabular}
\end{table}

@@

As we will see in Chapter \ref{chap:co-methylation}, this result is contradicted by my analyses of within-fragment co-methylation (__TODO__: Update once Chapter \ref{chap-comethylation} is updated.)

@@

Smoothing is particularly powerful for experiments with low sequencing coverage since the error in estimating the raw $\beta$-values is inversely proportional to the sequencing coverage and smoothing reduces this error. An important parameter when smoothing is the choice of window size over which to smooth, also known as the _bandwidth_ of the smoother. The default window size in `bsseq` is defined as one that contains at least 70 CpGs and is at least 2000 bp wide. `BiSeq` uses a much smaller default window size of 80 bp, a decision that is  driven by its focus on RRBS data which has a higher CpG density than whole-genome bisulfite-sequencing data. Ideally, a data-driven bandwidth would be used, one driven by the strength of co-methylation in the region, but this is difficult to implement in practice. A simpler but potentially useful hybrid would be to define different window sizes for genomic features with generally different co-methylation structures, e.g., CpG islands vs. the rest of the genome (see Chapter \ref{chap:co-methylation}) (__TODO: Revisit in light of updated results in Co-methylation chapter__).

>>>>

\begin{chapabstract}
Co-methylation is the dependence structure of DNA methylation data. This chapter critically reviews different definitions of co-methylation and puts them in the statistical framework of Chapter \ref{chap:wgbs_statistical_framework}. I highlight various shortcomings of existing methods for quantifying co-methylation, which led me to develop the methods and software described in  Chapter \ref{chap:co-methylation}. I also review how methods to detect differential methylation have sought to account for, and exploit, co-methylation.
\end{chapabstract}

@@

Let $0 \leq p_{h} \leq 1$ be the genome-wide average CpG methylation level for the sample from which $Z_{h, i}$ is sampled, that is, $Pr(Z_{h, i} = 1) = p_{h}$. For two distinct CpGs sequenced on two (potentially) distinct amplicons, $Z_{h, i}$ and  $Z_{h', i'}$, we want to compute the probability that the methylation states are identical, that is, $Pr(Z_{h, i} = Z_{h', i'})$. Not only may the two CpGs be from distinct amplicons, but in the original analysis the amplicons themselves may be from distinct samples with different genome-wide average methylation levels. This most general setting is described by $Z_{h, i} \eqd \text{Bernoulli}(p_{h})$ and $Z_{h', i'} \eqd \text{Bernoulli}(p_{h'})$, with $cor(Z_{h, i}, Z_{h', i'}) = c$. In this setting we can compute the theoretical $PIM^{\ast}$ as follows:

\begin{align*}
PIM^{\ast} &= Pr(Z_{h, i} = Z_{h', i'}) \\
        &= Pr(Z_{h, i} = 0, Z_{h', i'} = 0) + Pr(Z_{h, i} = 1, Z_{h', i'} = 1) \\
        &= Pr(Z_{h', i'} = 0 | Z_{h, i} = 0) Pr(Z_{h, i} = 0) +
           Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(Z_{h, i} = 1) \\
        &= \big [ (1 - p_{h'}) + c p_{h} \big ] (1 - p_{h}) +
           \big [ p_{h'} + c (1 - p_{h}) \big ] p_{h} \\
        &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
\end{align*}

@@

1. $PIM^{\ast}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'}$: Independent and non-identically distributed (_inid_), i.e., $Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h', i'} \eqd Bernoulli(p_{h'})$ and with $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.
2. $PIM^{\ast}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $Z_{h, i}, Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c \neq 0$.
3. $PIM^{\ast}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $Z_{h, i}, \allowbreak Z_{h', i'} \eqd Bernoulli(p)$ and $cor(Z_{h, i}, Z_{h', i'}) = c = 0$.

@@

The $167$ bp period "is similar to, but slightly shorter than, estimates of the average spacing of nucleosomes in plant chromatin" and Cokus et al. hypothesise that this periodicity is due to histones dictating access to DNA by the DNA methyltransferases.

@@

\citet{Ball:2009kh} use bisulfite padlock probes with $36$ bp single-end reads to profile CpG methylation at approximately $7,000$ CpGs in an Epstein-Barr virus transformed B-lymphocyte cell line. As part of their study, \citeauthor{Ball:2009kh} sought to investigate whether the co-methylation reported by \citet{Eckhardt:2006gh} occurs at the single-molecule level.

@@

\citet{Landan:2012kp} is notable for its use of the within-read information available from bisulfite-sequencing data. The authors explore the within-sample heterogeneity of methylation patterns at 4-tuples and 6-tuples of CpGs from ultra-deep bisulfite-sequencing ($> 10,000 \times$ median sequencing coverage of each CpG in the target) of 45 cancer-related CpG islands and study how methylation dynamics at this regions evolves over time. \citeauthor{Landan:2012kp} also use these data to estimate co-methylation.

As part of their analysis, \citeauthor{Landan:2012kp} construct "methylation linkage diagrams" based on a Pearson correlation estimate of within-fragment co-methylation. For each CpG 2-tuple in their targeted regions, they create a $2 \times 2$ contingency table. The general form of the table is shown below in Table \ref{tab:landan}. They then compute the Pearson correlation of each table[^cor_equivalence], $r = \frac{n_{mm} \times n_{uu} - n_{mu} \times n_{um}}{\sqrt{n_{u+} \times n_{m+} \times n_{+u} \times n_{+m}}}$. Importantly, $r$ is based on within-read measurements from a single sample and so these $r$ are estimates of within-read, within-sample co-methylation.

@@

\begin{table}[h]
\centering
\caption[Counts of reads at a CpG 2-tuple]{Summary of read counts at a CpG 2-tuple. For example, $n_{um}$ is the number of reads methylated at the first CpG and unmethylated at the second CpG. We use the 'plus' notation to denote the sum over the index it replaces, e.g., $n_{u+} = n_{uu} + n_{um}$.}
\label{tab:landan}
\begin{tabular}{@{}cccc@{}}
\cmidrule(l){2-4}
 & \multicolumn{3}{c}{Second CpG} \\
 \cmidrule(l){2-4}
\multirow{3}{*}{First CpG} & & & \\
 & $n_{uu}$       & $n_{um}$       & $n_{u+}$ \\
 & $n_{mu}$       & $n_{mm}$       & $n_{m+}$ \\ \cmidrule(l){2-4}
 & $n_{+u}$  & $\cdot n_{+m}$  & $n_{++} = n$      \\
\end{tabular}
\end{table}

@@

\begin{table}[h]
\centering
\caption[Notions of `subject' and `population' in methylation linkage]{Notions of `subject' and `population' in traditional genetic linkage and the "methylation linkage" of \citet{Landan:2012kp}. NB `person' could be any organism, e.g, `mouse', `dog', etc.}
\label{tab:linkage}
\begin{tabular}{@{}lll@{}}
\toprule
                    & Subject               & Population \\ \midrule
Genetic linkage     & Person                   & Population \\
Methylation linkage & (Haploid) genome of cell & Person     \\ \bottomrule
\end{tabular}
\end{table}

@@

As we will see in Chapter \ref{chap:co-methylation}, this result is contradicted by my analyses of within-fragment co-methylation using whole-genome data.

@@

Smoothing is particularly powerful for experiments with low sequencing coverage since the error in estimating the raw $\beta$-values is inversely proportional to the sequencing coverage and smoothing reduces this error. An important parameter when smoothing is the choice of window size over which to smooth, also known as the _bandwidth_ of the smoother. The default window size in `bsseq` is defined as one that contains at least 70 CpGs and is at least 2000 bp wide. `BiSeq` uses a much smaller default window size of 80 bp, a decision that is  driven by its focus on RRBS data which has a higher CpG density than whole-genome bisulfite-sequencing data. Ideally, a data-driven bandwidth would be used, one driven by the strength of co-methylation in the region, but this is difficult to implement in practice. A simpler but potentially useful hybrid would be to define different window sizes for genomic features with generally different co-methylation structures, e.g., CpG islands vs. the rest of the genome (see Chapter \ref{chap:co-methylation}).
